# ggseas R package
seasonal adjustment on the fly extension for ggplot2

Convenience functions that let you easily do seasonal adjustment on the fly with ggplot.

Depends on the [`seasonal` package](https://cran.r-project.org/web/packages/seasonal/index.html) to give you access to X13-SEATS-ARIMA.

## Installation
Install the stable version the usual way from CRAN:

```{r, eval = FALSE}
install.packages("ggseas")
```


or the latest version (bugs and all) from GitHub:
```{r, eval = FALSE}
devtools::install_github("ellisp/ggseas/pkg")
```


## Usage - seasonal adjustment on the fly
So far there are three types of seasonal adjustment possible to be incorporated
into a usual ggplot() command, substituting for where you'd normally have geom_line().

### X13-SEATS-ARIMA
```{r}
library(ggseas)
# make demo data
ap_df <- data.frame(
   x = as.numeric(time(AirPassengers)),
   y = as.numeric(AirPassengers)
)

# SEATS with defaults
ggplot(ap_df, aes(x = x, y = y)) +
   geom_line(colour = "grey80") +
   stat_seas(start = c(1949, 1), frequency = 12) +
   ggtitle("SEATS seasonal adjustment - international airline passengers") +
   ylab("International airline passengers per month")

# X11 with no outlier treatment
ggplot(ap_df, aes(x = x, y = y)) +
   geom_line(colour = "grey80") +
   stat_seas(start = c(1949, 1), frequency = 12, x13_params = list(x11 = "", outlier = NULL)) +
   ggtitle("X11 seasonal adjustment - international airline passengers") +
   ylab("International airline passengers per month")

ggplot(ldeaths_df, aes(x = YearMon, y = deaths, colour = sex)) +
   geom_point(colour = "grey50") +
   geom_line(colour = "grey50") +
   facet_wrap(~sex) +
   stat_seas(start = c(1974, 1), frequency = 12, size = 2) +
   ggtitle("Seasonally adjusted lung deaths in the UK 1974 - 1979") +
   ylab("Deaths") +
   xlab("(light grey shows original data;\ncoloured line is seasonally adjusted)") +
   theme(legend.position = "none")
```

### STL (LOESS-based decomposition)
```{r}
# periodic if fixed seasonality; doesn't work well:
ggplot(ap_df, aes(x = x, y = y)) +
   geom_line(colour = "grey80") +
   stat_stl(frequency = 12, s.window = "periodic")

# seasonality varies a bit over time, works better:
ggplot(ap_df, aes(x = x, y = y)) +
   geom_line(colour = "grey80") +
   stat_stl(frequency = 12, s.window = 7)
```

### Classical decomposition
```{r}
# default additive decomposition (doesn't work well in this case!):
ggplot(ap_df, aes(x = x, y = y)) +
   geom_line(colour = "grey80") +
   stat_decomp(frequency = 12)

# multiplicative decomposition, more appropriate:
ggplot(ap_df, aes(x = x, y = y)) +
   geom_line(colour = "grey80") +
   stat_decomp(frequency = 12, type = "multiplicative")
```

## Usage - seasonal decomposition on the fly
From version 0.2.0 I introduce a summary graphic decomposition, similar to what
you'd get with plot(stats::decompose(x)), but in the ggplot2 environment.  As well
as allowing ggplot2 look and feel of plots, you can also map a variable to the 
colour (or color) aesthetic, to allow two difference decompositions on the same
graphic.
```{r}
ggsdc(ap_df, aes(x = x, y = y), method = "decompose", frequency = 12) +
   geom_line()


ggsdc(ap_df, aes(x = x, y = y), method = "stl", frequency = 12, s.window = 7) +
   labs(x = "", y = "Air passenger numbers") +
   geom_point()

ggsdc(ldeaths_df, aes(x = YearMon, y = deaths, colour = sex), method = "seas", 
      frequency = 12, start = c(1949, 1)) +
      geom_line()
```

